---
title: "Performance of dfCount()"
author: "Dean Attali"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance of dfCount()}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(tidy = FALSE, comment = "#>")
```

## Performance of `dfCount()`

After some basic testing (not extremely thorough), I believe that `dfCount()`
performs much faster than its equivalent `table()` on large datasets, especially
when the data is numeric.  The analysis was done with the `microbenchmark`
package to compare the two functions on a few different datasets.

```{r dfCount-performance, results = "hide"}
library(rsalad)
library(dplyr)
library(microbenchmark)

# Prepare all the datasets to test on
fDat <- nycflights13::flights
largeIntDat <- data.frame(col = rep(1:25, 100000))
largeCharDat <- data.frame(col = rep(letters[1:25], 100000))
smallDat <- data.frame(col = rep(1:25, 100))

# Run the benchmarking
m <-
  microbenchmark(
    dfCount(fDat, "day"), table(fDat$day),
    dfCount(fDat, "dest"), table(fDat$dest),
    dfCount(largeIntDat, "col"), table(largeIntDat$col),
    dfCount(largeCharDat, "col"), table(largeCharDat$col),
    dfCount(smallDat, "col"), table(smallDat$col),
    times = 10
  )
```

```{r show-dfCount-performance, echo = FALSE, results = 'asis'}
knitr::kable(summary(m) %>% select(expr, min, mean, median, max, neval))
```

Every pair of rows corresponds to counting the same data using `dfCount()` vs
`table()`.  The results show that:

  - `dfCount()` was faster in all 4 large datasets
- `dfCount()` was an order of magnitude faster in both cases when the data
was numeric
- `dfCount()` was slower on very the small dataset


After performing this analysis, I've realized that the likely cause of the
speed boost is due to `dfCount()` relying on `dplyr`. After making that
realization, I found that `dplyr` also has a `count()` function, which
performs equally fast as `dfCount()`, which further supports the hypothesis
that the speed boost was thanks to `dplyr`. However, I still want to include
this function in the package because it took a lot of hard work (and
documentation!), and it also has a very differences from `dplyr::count()`. For
example, `dplyr::count()` does not sort by default, which I find to be
the less desired behaviour, and `dplyr::count()` does not have a
standard-evaluation version.
