---
title: "Package rsalad"
author: "Dean Attali"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package rsalad}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(tidy = FALSE, comment = "#>")
```

## Overview

`rsalad`, like any other salad, is a mixture of different healthy vegetables 
that you should be having frequently and that can make your life much better. 
Except that instead of vegetables, `rsalad` provides you with R functions.  

This package was born as a result of me constantly breaking the DRY principle
by copy-and-pasting functions from old projects into new ones.  Hence, the
functions in `rsalad` do not have a single common topic, but they are all 
either related to manipulating data.frames or general productivity utilities.

## Analysis

This vignette will introduce all the families of functions available in
`rsalad`, but will not dive too deeply into any one specific function.  To
demonstrate all the functionality, we will use the `nycflights13::flights`
dataset (information about ~335k flights departing from NYC) to visualize the
50 most common destinations of flights out of NYC.  While the analysis is not 
particularly exciting, it will show how to use `rsalad` proficiently.

### Load packages

Before beginning any analysis using `rsalad`, the first step is to load the
package. We'll also load `dplyr`, a package that every analysis workflow should use.

```{r load-lib, message = FALSE, warning = FALSE}
library(rsalad)
library(dplyr)
```

### Load data

First step is to load the `flights` dataset and have a peak at how it looks

```{r load-data, results = "hide"}
fDat <- nycflights13::flights
head(fDat)
```

```{r show-data, echo = FALSE, results = 'asis'}
knitr::kable(head(fDat))
```

### `%nin%` operator and `notIn()`

Let's say that for some reason we aren't interested in flights operated by
United Airlines (UA), Delta Airlines (DL) and American Airlines (AA). To choose
only carrier that are **not** part of that group, we can use the `%nin%`
operator, which is also aliased to `notIn()`.

```{r show-do-notIn, results = "hold"}
fDat2 <- fDat %>% filter(carrier %nin% c("UA", "DL", "AA"))
allCarriers <- fDat %>% select(carrier) %>% first %>% unique
myCarriers <- fDat2 %>% select(carrier) %>% first %>% unique

paste0("All carriers: ", paste(allCarriers, collapse = ", "))
paste0("My carriers: ", paste(myCarriers, collapse = ", "))
```

The `%nin%` operator is simply the negation of `%in%`, but can be a handy
shortcut. `lhs %nin% rhs` is equivalent to `notIn(lhs, rhs)`.  The following
code would have the same result as above:

```{r show-notIn-alias}
fDat2_2 <- fDat %>% filter(notIn(carrier, c("UA", "DL", "AA")))
identical(fDat2, fDat2_2)
```

For more information, see `?rsalad::notIn`.

### `move` functions: move columns to front/back

The `move` family of functions can be used to rearrange the column order of a
data.frame by moving specific columns to be the first (`moveFront()` and
`moveFront_()`) or last (`moveBack()` and `moveBack_()`) columns.  
The order in which the columns are passed in as arguments determines the order
in which the columns will be in the resulting data.frame, regardless of whether
the columns are moved to the front or back.  

These functions support non-standard evaulation (see function documentation
for more details).

For brevity, we will only keep a few columns in the data.

```{r select-cols, results = "hide"}
fDat3 <- fDat2 %>% select(carrier, flight, origin, dest)
head(fDat3)
```

```{r show-data-3, echo = FALSE, results = 'asis'}
knitr::kable(head(fDat3))
```

Now let's rearrange the columns to be in this order: dest, origin, carrier, 
flight.

```{r move, results = "hide"}
fDat4 <- fDat3 %>% moveFront(dest, origin)
head(fDat4)
```

```{r show-data-4, echo = FALSE, results = 'asis'}
knitr::kable(head(fDat4))
```

The same result can be achieved in different ways using other `move` functions.
```{r move-alias}
fDat4_2 <- fDat3 %>% moveFront_(c("dest", "origin"))
fDat4_3 <- fDat3 %>% moveBack(carrier, flight) %>% moveFront(dest)

all(identical(fDat4, fDat4_2), identical(fDat4, fDat4_3))
```

For more information, see `?rsalad::move`.

### `dfFactorize()`: convert data.frame columns to factors

Sometimes you want to convert all the character columns of a data.frame
into factors.  In our current data, we have three character variables
(dest, origin, carrier), but they all make more sense as factors.  Rather
than converting each column manually, we can use the `dfFactorize()` function.  

```{r do-dfFactorize}
str(fDat4)
fDat5 <- fDat4 %>% dfFactorize()
str(fDat5)
```

As you can see, calling `dfFactorize()` with no additional arguments converted
all potential factor columns into factors. Note that the integer column was
unaffected.  

By default, all character columns are coerced to factors, but we can also
specify which columns to convert or which columns to leave unaffected.

```{r dfFactorize-examples}
str(fDat4 %>% dfFactorize(only = "origin"))
str(fDat4 %>% dfFactorize(ignore = c("origin", "dest")))
```

For more information, see `?rsalad::dfFactorize`.

### `dfCount()`: count number of rows per group

Our goal is to see which destinations were the most common, so the next step
is to count how many observations we have for each destination.  This can be
achieved using the base R function `table()`:

```{r do-table}
head(table(fDat5$dest))
```

However, this is such a common task for me that I was not happy with the result
`table()` gives.  
Specifically:  

- `table()` returns a `table` object rather than the much more uesful
`data.frame`.  
- `table()` does not sort the resulting counts.  
- `table()` performs very slowly on large datasets, especially if the data is
numeric (see Performance section below).  

The `dfCount()` function provides an alternative way to count the data in a
data.frame column in an efficient way, sorts the results, and returns a
data.frame.  
Let's use dfCount to count the number of flights for each destination.

```{r do-dfCount, results = "hide"}
countDat <- fDat5 %>% dfCount("dest")
head(countDat)
```

```{r show-count-data, echo = FALSE, results = 'asis'}
knitr::kable(head(countDat))
```

Now our count data is in a nice data.frame format that can play nicely with
other data.frames, and can be easily merged/joined into the original dataset
if we wanted to.

Since we only want to see the 50 most common destinations, and the count data
is sorted in descending order, we can now easily retain only the 50 destinations
that appeared the most.

```{r keep-50}
countDat2 <- slice(countDat, 1:50)
```

For a performance analysis of `dfCount` vs `base::table`, see the
[dfCount performance vignette](https://github.com/daattali/rsalad/blob/master/vignettes/dfCountPerf.md).

For more information, see `?rsalad::dfCount`.

## Visual analysis

The [`ggExtra`](https://github.com/daattali/ggExtra) package has several
functions that can be used to plot the resulting data more efficiently. These
functions used to be part of this package, but are now in their own dedicated
package.

## Other functions

#### `spinMyR()`: create markdown/HTML reports from R scripts with no hassle
See the [spinMyR vignette](spinMyR.md) for information about this
function.


#### `tolowerfirst()`: convert first character to lower case

`rsalad` provides another function that can sometimes become handy.
`tolowerfirst()` can be used to convert the first letter of a string (or a 
vector of strings) into lower case. This can be useful, for example, when 
columns of a data.frame do not follow a consistent capitalization and you would 
like to lower-case all first letters.

```{r show-tolowerfirst}
df <- data.frame(StudentName = character(0), ExamGrade = numeric(0))
(colnames(df) <- tolowerfirst(colnames(df)))
```

For more information, see `?rsalad::tolowerfirst`.

#### `setdiffsym()`: symmetric set difference

When wanting to know the difference between two sets, the base R function
`setdiff()` unfortunately does not do exactly what you want because it is
*asymmetric*.  This means that the results depend on the order of the two
vectors passed in, which is often not the desired behaviour.  `setdiffsym`
implements symmetric set difference, whiich is a more intuitive set difference.

```{r show-setdiffsym}
setdiff(1:5, 2:4)
setdiff(2:4, 1:5)
setdiffsym(1:5, 2:4)
setdiffsym(2:4, 1:5)
```

For more information, see `?rsalad::setdiffsym`.

### `%btwn%` operator and `between()`

Determine if a numeric value is between the specified range. By default, the
range is inclusive of the endpoints.

```{r show-btwn}
5 %btwn% c(1, 10)
c(5, 20) %btwn% c(5, 10)
rsalad::between(5, c(5, 10))
rsalad::between(5, c(5, 10), inclusive = FALSE)
```

For more information, see `?rsalad::between`.
